Using Occupancy Theory and Hierarchical Models to Estimate Otter Population in Southeastern Minnesota

### 2010-02-28 Paper.tex
Added some edits to my sections.

    WinBUGS requires a representation of a model, data, and initial values on
    which a predetermined number of iterations of the MCMC algorithm are
    executed \cite{MacKenzie2006}. Because WinBUGS explicitly estimates missing
    data that is `padded out' by inserting NA values \cite{Spiegelhalter2003},
    we implement a nested indexing strategy to prevent this extraneous
    estimation on the actual data. This is a particularly important
    implementation detail in this case because the data we received from the
    Minnesota Department of Natural Resources was not consistent with respect to
    our covariates. Neither the number of observers nor number of observations
    after each snow event exhibited consistency. Our simulations, on the other
    hand, involve balanced data sets that are much less difficult to implement
    within WinBUGS and R.

@@

    \subsection{Our hierarchical model}
    For our model, we specified uniform \((0, 1)\) prior distributions for
    parameters \(\psi\) and \(\theta\), a common, natural choice given a lack of
    formal information about the prior or expert consensus \cite{MacKenzie2006}.
    We implemented models with normal and beta prior distributions on \(p\) and
    \(E\) and limited the range of \(p\) to the interval \((0.6, 1)\) and the
    range of \(E\) to the interval \((0, 0.4)\) in both models. We set the
    burn-in time for the MCMC simulation to 10,000 iterations. Each of these
    parameters are tracked and therefore are estimated. The covariates of this
    model are: observer, date, and the number of days since the last snow event.

@@

    To determine the efficacy of the standard model over the range of possible
    \(\psi\) and \(\theta\) values, we ran simulations with several combinations
    of \(\psi\) and \(\theta\), ran the model on each simulation, and compared
    the model's estimates to the simulation's parameters. A table detailing the
    model's performance lies below:

@@

    Generally, the model is most effective for lower values of \(\psi\) and
    \(\theta\) but will either converge well or provide useful estimates for the
    vast majority of combinations of \(\psi\) and \(\theta\) values.

@@

    \subsection{Results}
    Below, in tables, lie our estimates for all of the Minnesota DNR's data from
    2003 and 2004. Below each table is detailed commentary about the ability of
    the standard hierarchical model to converge, given the corresponding data.

>>>>

    WinBUGS requires a representation of a model, data, and initial values on
    which a predetermined number of iterations of the MCMC algorithm are
    executed \cite{MacKenzie2006}.
    The data we received from the
    Minnesota Department of Natural Resources was not fixed with respect to
    our covariates and required us to implement one of the three common
    strategies used to deal with irregular datasets \cite{Spiegelhalter2003}.
    Our simulations, on the other
    hand, ran on datasets with a fixed number of observers, snow events, and
    observations after each snow event, which are much easier to implement
    within WinBUGS and R.

@@

    \subsection{Our hierarchical model}
    For our model, we specified noninformative uniform \((0, 1)\) prior
    distributions for
    parameters \(\psi\) and \(\theta\), a common, natural choice given a lack of
    formal information about the prior or expert consensus \cite{MacKenzie2006}.
    We implemented models with normal and beta prior distributions on \(p\) and
    \(E\) and limited the range of \(p\) to the interval \((0.6, 1)\) and the
    range of \(E\) to the interval \((0, 0.4)\) in both models. We set the
    burn-in time for the MCMC simulation to 10,000 iterations. Each of these
    parameters are tracked and therefore are estimated. The covariates of this
    model are: observer, date, and the number of days since the last snow event.
    It was necessary to preprocess the Minnesota DNR's data to account for its
    irregularities.

@@

    To determine the efficacy of the standard model over the range of possible
    \(\psi\) and \(\theta\) values, we ran simulations with several combinations
    of \(\psi\) and \(\theta\), ran the model on each simulation, and compared
    the model's estimates to the simulation's parameters. The model with beta
    prior distributions on \(p\) and \(E\) performed better than the model with
    normal prior distributions on the same parameters. A table detailing the
    model's performance with the beta prior distributions lies below:

@@

    Generally, the model is most effective for lower values of \(\psi\) and
    \(\theta\) but will either converge well or provide useful estimates for the
    vast majority of combinations of \(\psi\) and \(\theta\) values. 21 of the
    model's 25 confidence intervals for \(\psi\) include its actual value. With
    respect to \(p\), all of the model's confidence intervals for each of the
    three values of \(p\) include its actual value for all \(\psi\) and
    \(\theta\) values. On the other hand, the model accurately estimates \(E\)
    for low values of \(\theta\), but consistently overestimates \(E\) when
    \(\psi\geq0.4\) and \(\theta\geq0.6\) regardless of \(E\)'s actual value.

@@

    \subsection{Our estimates}
    Below, in tables, lie our estimates for all of the Minnesota DNR's data from
    2003 and 2004. Below each table is detailed commentary about the ability of
    the standard hierarchical model to converge, given the corresponding data.
